---
title: "blockr.stats: statistical blocks and their engine functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{blockr.stats: statistical blocks and their engine functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(blockr.stats)
```

## What this package is

`blockr.stats` extends [blockr.core](https://github.com/blockr-org/blockr.core)
with no-code statistical blocks. Every block does one thing and every block
ends in the same place: a **tidy data frame** that blockr.core's generic
renderers (drilldown table, coefficient plot, scatter, step curve) draw
without any block-specific plotting code.

The blocks cover an applied-statistics-basics *spine*

* **Describe** -- per-variable summaries, frequency / cross tabulation
* **Compare** -- adaptive significance tests, p-value adjustment, effect sizes
* **Associate** -- pairwise correlation matrix
* **Regress** -- `lm` / `glm` / `aov` model fitting and a generic broom adapter
* **Nonparametric** -- folded into the adaptive test stack

plus one **Advanced** tier -- survival (Kaplan-Meier, Cox, competing risks) --
deliberately kept as a single separable cluster so it can later move to a
dedicated domain package.

This vignette walks through the *engine functions* behind the blocks. They are
plain functions with no Shiny dependency, so you can use them directly, test
them, and read them to understand what each block emits. In a board you would
not call them by hand -- you would wire the corresponding `new_*_block()` into
a blockr pipeline -- but the data frames are identical either way.

## Architecture: model objects flow, broom adapts, renderers draw

The key design decision is the **renderer boundary**. Two kinds of block exist:

1. **Model-fitting blocks** (`new_model_block()`, `new_lm_block()`,
   `new_survival_block()`) emit a *fitted model object* (an `lm`, `glm`,
   `coxph`, `survfit`, `cuminc`). Their own preview is a generic `summary()`.
2. **Tidying blocks** -- chiefly the **broom adapter**
   (`new_broom_block()`) -- consume a fitted model upstream and emit a tidy
   data frame.

Everything that produces a table directly (`describe_numeric()`,
`correlation_matrix()`, `effect_size()`, the adaptive tests) skips step 1 and
emits tidy frames straight away. The result: no block carries plotting code.
A coefficient plot is just the generic drilldown reading `estimate`,
`conf.low`, `conf.high`; a survival curve is the generic step renderer reading
`time`, `estimate`, `group`.

```
 data ─▶ [model block] ─▶ fitted object ─▶ [broom adapter] ─▶ tidy df ─▶ renderer
 data ─▶ [describe / correlate / test / effect-size] ───────▶ tidy df ─▶ renderer
```

## Describe

`describe_numeric()` returns one row per numeric variable -- `n`, `n_missing`,
`mean`, `sd`, `median`, `q1`, `q3`, `min`, `max` -- in base R, no easystats
dependency.

```{r describe}
describe_numeric(mtcars, vars = c("mpg", "hp", "wt"))
```

With `vars = NULL` (the default) it summarises every numeric column. This is
the frame the **Descriptives** block emits.

## Associate

`correlation_matrix()` is the one matrix-valued step base R has no tidy verb
for, so the function emits it *long* -- `var_x`, `var_y`, `r` -- which is
exactly what a drilldown table or diverging-colour heatmap wants.

```{r correlation}
cm <- correlation_matrix(mtcars[, c("mpg", "hp", "wt", "qsec")],
                         method = "pearson")
head(cm, 8)
```

`method` accepts `"pearson"`, `"spearman"`, or `"kendall"`; `max_cols` caps
the number of numeric columns so a wide frame cannot blow up into a huge
pairwise product.

## Regress: a model object, then the broom adapter

Fit an ordinary model the normal way, then hand it to `broom_apply()` -- the
single generic adapter that turns *any* fitted model into a tidy frame for the
renderers.

```{r broom-tidy}
fit <- lm(mpg ~ wt + hp, data = mtcars)

broom_apply(fit, output = "tidy", conf_int = TRUE)
```

`output = "tidy"` is the coefficient table (with `conf.low` / `conf.high` so
the generic drilldown can draw a coefficient plot). The other two outputs:

```{r broom-glance}
broom_apply(fit, output = "glance")
```

```{r broom-augment}
head(broom_apply(fit, output = "augment"))
```

With `qq = TRUE`, `augment` additionally gets `.qq_theoretical` /
`.qq_sample`, so a QQ plot becomes a plain scatter -- again, no plotting code
in the block:

```{r broom-qq}
head(broom_apply(fit, output = "augment", qq = TRUE)[,
     c(".resid", ".qq_theoretical", ".qq_sample")])
```

This is why a single `new_broom_block()` serves `lm`, `glm`, `aov` and Cox
models alike -- the renderer only ever sees a tidy frame.

## Compare: effect sizes

`effect_size()` consumes a fitted `lm`/`aov` and emits a tidy effect-size
frame -- `term`, `measure`, `estimate`, `conf.low`, `conf.high` -- with **no
interpretation labels** (the package does not editorialise "small/medium/
large").

```{r effect-size}
aov_fit <- aov(mpg ~ factor(cyl), data = mtcars)

effect_size(aov_fit, measure = "partial_eta2")
```

`measure` is one of `"eta2"`, `"partial_eta2"`, `"omega2"`, `"r2"`, `"d"`,
`"g"`. The partial-eta-squared CI is obtained by inverting the noncentral F
(`es_ncp_ci()`), returning `NA` bounds on non-convergence rather than
erroring.

## Advanced: survival

`fit_survival()` is the Advanced-tier engine -- the single extraction seam to
a future domain package. It returns the *fitted object* (`survfit`, `coxph`,
or `cuminc`); the broom adapter tidies it downstream exactly like any other
model.

```{r survival-km}
lung <- survival::lung
# the helper expects event = 1, censor = 0; lung codes 2 = death, 1 = censor
lung$dead <- lung$status - 1L

km <- fit_survival(lung, type = "km",
                   time_var = "time", event_var = "dead",
                   group_var = "sex")
class(km)
```

`tidy()` (broom's `survfit` method) turns the curve into long
`time` / `estimate` / `strata` points -- a drilldown line+step:

```{r survival-tidy}
km_tidy <- broom::tidy(km)
head(km_tidy[, c("time", "estimate", "n.risk", "strata")])
```

A Cox model flows the same way -- `fit_survival(type = "cox")` returns a
`coxph`, and `broom_apply(cox, "tidy")` yields hazard ratios with CIs (a
coefficient plot via the same generic renderer). Competing-risks
(`type = "cif"`) returns a `cmprsk::cuminc` object; broom has no method for
it, so the package ships `tidy.cuminc()`, emitting long
`group` / `time` / `estimate` so cumulative-incidence curves render with the
same step renderer as Kaplan-Meier.

## Putting it together in a board

In an interactive blockr board you would not call these functions directly.
You would compose the registered blocks -- e.g. a data block, then
`new_model_block()`, then `new_broom_block()`, then a generic plot block --
and blockr.core would wire the reactive graph. The functions in this vignette
are the *contract*: whatever a block produces is one of the frames shown
above, which is what makes a single set of generic renderers enough for the
whole package.

```{r registry}
# the blocks this package registers
ls("package:blockr.stats", pattern = "^new_.*_block$")
```
