---
title: "blockr.stats: statistical blocks and their engine functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{blockr.stats: statistical blocks and their engine functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(blockr.stats)
```

## What this package is

`blockr.stats` extends [blockr.core](https://github.com/blockr-org/blockr.core)
with no-code statistical blocks. Every block does one thing and every block
ends in the same place: a **tidy data frame** that blockr.core's generic
renderers (drilldown table, coefficient plot, scatter, step curve) draw
without any block-specific plotting code.

There are six blocks:

* **Model** -- `lm` / `glm` fitting from a formula
* **Model summary** -- a fitted model rendered as a card: model facts line
  plus a coefficient table with an inline forest. It calls `broom::tidy()` and
  `broom::glance()` itself, so it needs no adapter in between, and its own
  value is the tidy coefficient frame
* **Broom adapter** -- any fitted model to a tidy frame
* **Statistical test** -- one adaptive block covering the `stats` test family
* **Correlate** -- pairwise correlation matrix, rendered as a heatmap table
* **Survival** -- Kaplan-Meier, Cox, competing risks, kept as a single
  separable cluster so it can later move to a dedicated domain package

Descriptive tables (per-variable summaries, counts, crosstabs) are
deliberately absent: `blockr.viz::new_summary_table_block()` already produces
them for mixed types with by-group columns, and blockr.dplyr covers the rest.

Every block emits **standard R** -- `stats::lm()`, `survival::survfit()`,
`broom::tidy()`, and so on -- so a generated pipeline reads as ordinary code
you could run without blockr loaded.

## Architecture: model objects flow, broom adapts, renderers draw

The key design decision is the **renderer boundary**. Two kinds of block exist:

1. **Model-fitting blocks** (`new_model_block()`, `new_survival_block()`)
   emit a *fitted model object* (an `lm`, `glm`,
   `coxph`, `survfit`, `cuminc`). Their own preview is a generic `summary()`.
2. **Tidying blocks** -- the **model summary**
   (`new_model_summary_block()`) and the **broom adapter**
   (`new_broom_block()`) -- consume a fitted model upstream and emit a tidy
   data frame. The summary block also draws that frame as a card; the adapter
   hands it on untouched, and adds `augment()` for per-observation output.

The blocks that produce a table directly (the adaptive tests, correlate) skip
step 1 and emit tidy frames straight away. The result: no block carries
plotting code. A coefficient plot is just the generic drilldown reading
`estimate`, `conf.low`, `conf.high`; a survival curve is the generic step
renderer reading `time`, `estimate`, `group`.

```
 data ─▶ [model block] ─▶ fitted object ─▶ [broom adapter] ─▶ tidy df ─▶ renderer
 data ─▶ [test / correlate] ───────────────────────────────▶ tidy df ─▶ renderer
```

## Regress: a model object, then the broom adapter

Fit an ordinary model the normal way; the broom adapter block then emits
**standard `broom` code** -- `broom::tidy()` / `glance()` / `augment()` -- to
turn *any* fitted model into a tidy frame for the renderers.

```{r broom-tidy}
fit <- lm(mpg ~ wt + hp, data = mtcars)

broom::tidy(fit, conf.int = TRUE)
```

`output = "tidy"` is the coefficient table (with `conf.low` / `conf.high` so
the generic drilldown can draw a coefficient plot). The other two outputs:

```{r broom-glance}
broom::glance(fit)
```

```{r broom-augment}
head(broom::augment(fit))
```

The block inlines a couple of conveniences into the generated expression
without leaving standard R: when QQ columns are switched on, `augment` gains
`.qq_theoretical` / `.qq_sample` (from `stats::qqnorm()`) so a QQ plot becomes
a plain scatter; and the `tidy` frame gets model-aware column `label`
attributes the renderers display. This is why a single `new_broom_block()`
serves `lm`, `glm`, `aov`, Cox and competing-risks alike -- the renderer only
ever sees a tidy frame.

## Advanced: survival

The survival block emits **standard R** -- `survival::survfit()` / `coxph()`
for Kaplan-Meier and Cox, and an inline `cmprsk::cuminc()` for competing
risks. The broom adapter tidies the fitted object downstream like any other
model.

```{r survival-km}
lung <- survival::lung
# event = 1, censor = 0; lung codes 2 = death, 1 = censor
lung$dead <- lung$status - 1L

km <- survival::survfit(survival::Surv(time, dead) ~ sex, data = lung)
class(km)
```

`tidy()` (broom's `survfit` method) turns the curve into long
`time` / `estimate` / `strata` points -- a drilldown line+step:

```{r survival-tidy}
km_tidy <- broom::tidy(km)
head(km_tidy[, c("time", "estimate", "n.risk", "strata")])
```

A Cox model flows the same way -- `survival::coxph(...)` returns a `coxph`, and
the broom adapter's `tidy` yields hazard ratios with CIs (a coefficient plot
via the same generic renderer). Competing-risks emits `cmprsk::cuminc(...)`;
broom has no method for it, so the package ships `tidy.cuminc()`, emitting long
`group` / `time` / `estimate` so cumulative-incidence curves render with the
same step renderer as Kaplan-Meier.

## Putting it together in a board

In an interactive blockr board you would not call these functions directly.
You would compose the registered blocks -- e.g. a data block, then
`new_model_block()`, then `new_broom_block()`, then a generic plot block --
and blockr.core would wire the reactive graph. The functions in this vignette
are the *contract*: whatever a block produces is one of the frames shown
above, which is what makes a single set of generic renderers enough for the
whole package.

```{r registry}
# the blocks this package registers
ls("package:blockr.stats", pattern = "^new_.*_block$")
```
